---
output:
  html_document:
    variant: markdown_github
    keep_md: true
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.retina = 2,
  out.width = "75%",
  dpi = 96
)
knit_hooks$set(pngquant = hook_pngquant)
```

# spectralGraphTopology
Learns the topology of a class of graphs such as K-component, tree, bipartite,
and regular.

See vignette for a detailed documentation with several illustrative examples.

The package is based on the paper:

## Installation
```{r, eval = FALSE}
# Installation from GitHub
install.packages("devtools")
devtools::install_github("dppalomar/spectralGraphTopology")
```

Alternatively, one can install the development version as follows:
```
$ git clone https://github.com/dppalomar/spectralGraphTopology.git
$ cd spectralGraphTopology
$ make build && make install
```

```{r, eval = FALSE}
# Getting help
library(spectralGraphTopology)
help(package = "spectralGraphTopology")
package?spectralGraphTopology
?spectralGraphTopology

# Citing this work
citation("spectralGraphTopology")
```


## Vignette
For more detailed information, please check the vignette:
[GitHub-html-vignette](https://rawgit.com/dppalomar/spectralGraphTopology/master/vignettes/SpectralGraphTopology-vignette.html),
[GitHub-pdf-vignette](https://rawgit.com/dppalomar/spectralGraphTopology/master/vignettes/SpectralGraphTopology-vignette.pdf),


## Usage of `learnGraphTopology()`


## Links
Package: [GitHub](https://github.com/dppalomar/spectralGraphTopology)
README file: [GitHub-readme](https://rawgit.com/dppalomar/spectralGraphTopology/master/README.html)
